Patterns of stress tolerance versus resistance in the thermal reactionomes of cold and warm-climate ant species

### 2014-08-01 doc/ApTranscriptome_manuscript.Rmd
manuscript v-1.2 sent to co-authors

---
title: "Patterns of stress tolerance versus resistance in the thermal reactionomes from populations of cold and warm-climate ant species"
author: \centering{John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1}, James Vincent \textsuperscript{2}, Mahesh Vangala \textsuperscript{2}, Robert R. Dunn \textsuperscript{3}, Nathan J. Sanders \textsuperscript{4}, Nicholas J. Gotelli \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1} \\ \vspace{5 mm} \textsuperscript{1} Department of Biology, University of Vermont \\ \textsuperscript{2} Vermont Genetics Network, University of Vermont \\ \textsuperscript{3} Department of Biological Sciences, North Carolina State University \\ \textsuperscript{4} Center for Macroecology, Evolution and Climate, University of Copenhagen}
csl: ecology.csl
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  html_document: null
  word_document: default
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

**250 words allowed - need to cut 56**

Thermal tolerance is a key trait governing the distributions of species and underlies their future responses to climate change. In this study, we demonstrate how insight on the evolution of thermal tolerance can be gained through characterizing an organism’s thermal “reactionome” – the reaction norm for all genes in an organism's transcriptome measured across a thermal gradient. We characterized the reactionomes of two populations of parapatric ant species of eastern U.S. deciduous forests: the northern cold-climate Aphaenogaster rudis and the southern warm-climate Aphaenogaster carolinensis . After exposing laboratory-acclimated individuals to a range of realistic temperatures (0° - 38.5°C), we found that 10% of genes were thermally responsive. When the two species were compared, the warm-adapted *A. carolinensis* displayed higher basal levels of gene expression, \todo{check this statement} a lower magnitude of upregulation at thermal extremes, upregulated of fewer heat shock proteins, and greater inertia of changes in gene expression in response to thermal extremes. This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability. Moreover, a surprisingly large proportion of genes showed divergent patterns of response to temperature between the two species, suggesting that they have different strategies for coping with temperature stress. *Aphaenogaster carolinensis* shut down more genes than *A. picea* in response to thermal extremes, whereas the same genes tended to be upregulated in *A. picea*. These included genes functionally involved in \todo{check} metabolism, cellular organization and repair. Altogether, these results suggest that *A. picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas *A. carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress. More broadly, this study adds to a growing body of evidence that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention, despite enzyme efficiency tradeoffs that such mechanisms entail.

@@

Here, we present a novel method for RNAseq analysis: the *thermal reactionome*, which is a characterization of the reaction norm for all genes in an organism's transcriptome across a thermal gradient. Rather than focusing on the identity of individual genes with differential expression, we compare overall patterns in the reactionomes of two closely-related temperate-zone species to gain insight into how spatially divergent selection has changed patterns of thermally-responsive gene expression. We apply this approach in a comparative study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996], which have more southern and northern distributions, respectively, and diverged approximately .... years ago (bernice??). Temperature is a potentially strong selective force on ground-nesting *Aphaenogaster* colonies, which must tolerate seasonally freezing winters and hot summers. On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@lubertazzi2012].

*A. carolinensis* lives in a climate that is hotter and has a longer growing season than that of *A. picea* (Supplemental Information). Thus, patterns of gene expression underlying tolerance of thermal extremes is likely to differ between these species in a number of ways. First, the number of genes showing plasticity of expression to thermal extremes may differ between the species. Finding more genes are up-regulated to high temperatures in the colder-climate species *A. picea* would indicate that this species relies on activation of the cellular stress response [@kultz2005] and other protective genes to a greater extent than *A. carolinensis*, while the opposite would be true for cold temperatures. Conversely, finding that more genes are down-regulated in response to thermal extremes would indicate that a species relies to a greater extent upon proteome stability [@ghosh2010] and behavioral mechanisms [@sunday2014] to tolerate thermal stress.

@@

Importantly, none of the above hypotheses depend on the functional annotation of the responsive genes, which can be inaccurate in non-model systems [@primmer2013; @skunca2012]. However, insight on the molecular toolkit involved in adaptation to thermal extremes can be gained through circumspect examination of the functional annotation of responsive genes. For example, we expect genes in the cellular stress response [@kultz2005] to be up-regulated in response to thermal extremes, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated in response to thermal extremes [@somero2011].

@@

The Trinity *de novo* transcriptome assembly included 126,172 genes with a total length of 100 million bp. Filtering to remove redundant or chimeric reads resulted in an assembly with 105,536 genes. The total length was 63 million bp with an N~50~ length of 895 bp and a mean gene size of 593 bp. Of the 105,536 filtered genes, 55,432 had hits to the NCBI-nr database. Of these, 38,711 genes mapped to GO terms, 1,659 genes were identified to an enzyme and 18,935 genes mapped to a domain with >50% coverage. We removed 5,675 genes identified as known contaminants, leaving 99,861 clean transcripts. Of these, 47% have best hits to genes from the class Insecta [(Supplementary Krona plot)](http://johnstantongeddes.org/assets/files/ApTranscriptome_FA.html).

```{r mosaic_plot, echo=FALSE, results='asis', fig.cap = "Mosaic plot of the number of thermally-responsive genes by expression type. The width of the columns is proportional to the number of genes for each expression type for *A. carolinensis* while the height of the rows is proportional to the number of genes for each expression type for *A. picea.* The rows are colored by expression type for *A. picea*. The number of genes are labeled in two boxes for scale."}
mosaic.colors <- brewer.pal(5, "Blues")
par(mar = c(5.1, 3.1, 2.1, 7.5))
mosaicplot(tt2, ylab = expression(italic("A. picea")), xlab = "", main = "", color = mosaic.colors, cex.axis = 0.01)
labels <- c("Low", "Intermediate", "High", "Bimodal", "NotResp")
text(c(.1,.5,.75,.85,.99), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE, font = 1)
text(.5, par("usr")[4]+.1, "A. carolinensis", xpd = TRUE, font = 3)
text(.15, .7, "2,594")
text(.85, .03, "223", cex = .75)
legend(1.05, .9, fill = mosaic.colors, c("Low", "Intermediate", "High", "Bimodal", "NotResp"), xpd=TRUE)
```

We identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` genes with expression patterns that are thermally-responsive at 5% FDR. Thus, at a minimum `r round(9100/99861, 2)*100`% of genes are thermally-responsive. However, the closest related ant species with sequenced genomes have only about 17,00 genes [@gadau2012], and assuming the number is similar for *Aphaenogaster*, this suggests that up to 50% of the genome could be thermally-responsive. The reaction norms for each responsive gene can be searched by functional annotation or GO category and viewed interactively at [https://johnsg.shinyapps.io/ApRxN-shinyapp/].

Of the thermally-responsive genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` have an expression pattern that differs between the two species. These differences were not only in magnitude, as the number of transcripts in each expression category differed significantly between the species ($\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig. 1). In both species, the majority of genes were in the **Low** category (Fig. 1). For *A. carolinensis*, this was followed by **Intermediate**, **High** and then **Bimodal**, whereas for *A. picea* the second largest category was **Bimodal** followed by **High** and finally **Intermediate** (Fig. 1). There were `r colSums(tt2)[which(colnames(tt2) == "NotResp")]` genes that were responsive in *A. carolinensis* but not *A. picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` that were responsive in *A. picea* but not *A. carolinensis*.

## Species comparisons

To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the temperature of gene activation for **High** and **Low** genes, and the variance in expression for the **Intermediate** and **Bimodal** genes. The **High** genes in *A. carolinensis* were activated on average at `r T_on.high.ttest$estimate[1]`°C compared to `r T_on.high.ttest$estimate[2]`°C for *A. picea* (*t*~df=`r T_on.high.ttest$parameter`~ = `r T_on.high.ttest$statistic`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For genes with increased expression at **Low** temperatures, there was no difference in the mean temperature of gene activation (*t*~df=`r T_on.low.ttest$parameter`~ = `r T_on.low.ttest$statistic`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2). For **Bimodal** expressed genes, we found no difference between species in the variance of the reaction norms (*t*~df=`r round(t.varbim$parameter, 0)`~ = `r round(t.varbim$statistic, 2)`, *P* < `r round(t.varbim$p.value, 2)`). In contrast, **Intermediate** genes from *A. carolinensis* had a broader thermal range of stable gene expression than those from *A. picea* (*t*~df=`r round(t.varint$parameter, 0)`~ = `r round(t.varint$statistic, 2)`, *P* < `r max(t.varint$p.value, 0.001)`, 95% CI difference `r paste(round(t.varint$conf.int[1], 3), "--", round(t.varint$conf.int[2], 3))`).

@@

We looked for evidence of genetic assimilation by comparing gene expression at intermediate temperatures for genes that were up-regulated at low temperatures in *A. carolinensis* or high temperatures in *A. picea*. Contrary to expectations, for **Low** genes in *A. carolinensis* gene expression at intermediate temperatures was lower in *A. picea* than *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`) . Moreover, for **High** genes in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`) and **Intermediate** genes in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`), gene expression at intermediate temperatures was always higher in *A. carolinensis*.

In addition to shifts in position, we quantified the differences between species in the mean and shape of the reaction norms of thermally-responsive genes. Species differed by mean expression (*t*~df=`r t.mean$parameter`~ = `r t.mean$statistic`, *P* < `r max(t.mean$p.value, 0.001)`) which was about `r round(exp(t.mean$estimate))` TPM greater in *A. carolinensis* than *A picea*, and mean slope (*t*~df=`r t.slope$parameter`~ = `r t.slope$statistic`, *P* < `r max(t.slope$p.value, 0.001)`), which was about `r round(exp(t.mean$estimate))` $\Delta$TPM/degree greater in *A. picea*. Curvature (*t*~df=`r t.curvature$parameter`~ = `r t.curvature$statistic`, *P* < `r round(max(t.curvature$p.value, 0.001), 3)`) and wiggle (*t*~df=`r t.wiggle$parameter`~ = `r t.wiggle$statistic`, *P* < `r max(t.wiggle$p.value, 0.001)`) did not differ between species. Of the total differences ($\Delta T$) in reaction norms between species for each gene, about `r round(qlmean[2] * 100, 0)`% (95% CI `r paste(round(qlmean[1] * 100, 0), "--", round(qlmean[3] * 100, 0))`%) and `r round(qlslope[2] * 100, 0)`% (95% CI `r paste(round(qlslope[1] * 100, 0), "--", round(qlslope[3] * 100, 0))`%) were due to shifts in mean and slope, respectively.

@@

As the molecular pathways involved in thermal response to cold and hot temperatures differ, we performed gene set enrichment analysis on each expression category separately. First, we looked at the genes shared in each expression category (e.g. the intersection) for each species (Table S1). Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched (*P* < 0.01) terms for **High** genes. Clustering of terms by semantic similarity yielded 3 clusters, with 5 of the 7 terms being involved in "regulation of biosynthetic/metabolic process". For **Low** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` GO terms were enriched. These terms grouped into 15 clusters, including "protein modification", "protein metabolic process", "lipid metabolic process", "regulation of signaling", "protein transport/localization", "detection of stimulus", and "chondroitin sulfate biosynthesis". For **Bimodal** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched. These terms grouped into 5 clusters, including "biological regulation", "regulation of cellular biosynthesis", "gene expression", "signaling", and "protein modification process". For **Intermediate** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched. These terms grouped into 3 clusters, including "lipid biosynthesis/metabolism", "glycoprotein biosynthesis/metabolism", and "DNA organization". Overall, GSEA indicates that molecular responses to thermal extremes involve large-scale cellular and organizational changes, though the molecular responses to high temperatures are more limited than those to low temperatures.

We further performed GSEA for the genes that shifted from being shut-down at extreme temperatures, **Intermediate**, in *A. carolinensis* to having increased expression in **Low** temperatures in *A. picea*. This group included `r nrow(Ar.int.A22.low.gsea)` GO terms that grouped into 12 clusters, including "protein metabolic process", "glycosylation" "transport", "lipid metabolic process", "GTP metabolism", "positive regulation of cellular process" and "organelle organization". For **Intermediate** genes in *A. carolinensis* that shifted **Bimodal** in *A. picea*, there were `r nrow(Ar.int.A22.bim.gsea)` enriched GO terms including "phospholipid catabolic process", "L-amino acid transport", "macromolecule metabolic process", "gluatmate receptor signaling pathway" and "histone modification". These results indicate that in response to thermal extremes, the warmer-climate species, *A. carolinensis*, shuts down metabolism to a greater extent than the colder-climate species, *A. picea*, which increases metabolism.

@@

Thermal tolerance is a key trait for determining the survival of individual organisms, the persistence of populations and the geographic limits of species. In this study, we applied a novel regression design to characterize the thermal reactionome of parapatric ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis*, found in colder and warmer environments, respectively, that are known to differ in their thermal tolerances [@warren2013]. We found that at least 10%, and potentially 50%, of the genes in these species were thermally-responsive.

The patterns of gene expression between species differed substantially, with more down-regulation in *A. carolinensis* (**Intermediate** genes) and more up-regulation of gene expression in *A. picea* in response to thermal extremes (**High** and **Low** genes; Fig. 1, 3). The results of the gene set enrichment analysis give an indication of what might drive this pattern. The genes that are up-regulated at cold temperatures (**Low**) in *A. picea* but shut-down **Intermediate** by *A. carolinensis* include metabolism, transport and regulation GO clusters. Thus, *A. carolinensis* shuts-down metabolism to a lesser extent than *A. picea* and relies on other mechanisms such as behavior or proteome stability to tolerate thermal stress. In contrast, *A. picea* increases metabolism to a greater extent than *A. carolinensis*, potentially to mount a greater molecular response to thermal stress. Consistent with this hypothesis, we find that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A. picea*, but only 4 do in *A. carolinensis*. This result is consistent with work in corals [@barshis2013] that shows that organisms in more thermally-stressful environments are more likely to invest in a tolerance strategy, while organisms from less stressful environments rely on plastically induced resistance.

@@

Given the differences in the number of thermally-responsive genes between the two species (Fig. 1), a striking pattern is that 3-4x more genes are upregulated at cold than high temperatures in both species (Fig. 3). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures are more constrained than critical minimum temperatures. Moreover, in an experiment with 94 species of *Drosophila*, Kellermann et al [-@kellermann2012] showed that limited genetic variation among species (e.g. phylogenetic inertia) rather than similar selection pressures underlie greater similarity of upper than lower thermal limits. Thus, when coupled with phenotypic data, our results demonstrate that transcriptomic studies can provide insight on evolutionary potential. For *Aphaenogaster*, our results suggest that adaptability to increased temperatures is limited, and these species risk replacement by more heat-tolerant competitors [@rodriguez-cabal2012].

>>>>

---
title: "Patterns of stress tolerance versus resistance in the thermal reactionomes of cold and warm-climate ant species"
author: \centering{John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1},
  James Vincent \textsuperscript{2}, Mahesh Vangala \textsuperscript{2}, Robert R.
  Dunn \textsuperscript{3}, Nathan J. Sanders \textsuperscript{4}, Nicholas J. Gotelli
  \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1} \\ \vspace{5 mm} \textsuperscript{1}
  Department of Biology, University of Vermont \\ \textsuperscript{2} Vermont Genetics
  Network, University of Vermont \\ \textsuperscript{3} Department of Biological Sciences,
  North Carolina State University \\ \textsuperscript{4} Center for Macroecology,
  Evolution and Climate, University of Copenhagen}
csl: ecology.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  html_document: null
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

**250 words allowed - need to cut 26**

Thermal tolerance is a key trait governing the distributions of species and underlies their future responses to climate change. In this study, we demonstrate how insight on the evolution of thermal tolerance can be gained through characterizing an organism’s thermal “reactionome” – the reaction norm for all genes in an organism's transcriptome measured across a thermal gradient. We characterized the reactionomes of two populations of ant species of eastern U.S. forests: the northern cold-climate Aphaenogaster picea and the southern warm-climate Aphaenogaster carolinensis . We found that  at least 10% of genes were thermally responsive. When the two species were compared, the warm-adapted *A. carolinensis* displayed higher basal levels of gene expression,, upregulated of fewer heat shock proteins, and had greater inertia of changes in gene expression in response to thermal extremes. This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability. A surprisingly large proportion of genes showed divergent patterns of response to temperature between the two species, suggesting that they have different strategies for coping with temperature stress. *Aphaenogaster carolinensis* shut down more genes than *A. picea* in response to thermal extremes, whereas the same genes tended to be upregulated in *A. picea*. These included genes functionally involved in metabolism, cellular organization and repair. Altogether, these results suggest that *A. picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas *A. carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress. More broadly, this study adds to a growing body of evidence that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.

@@

Here, we present a novel method for RNAseq analysis: the *thermal reactionome*, which is a characterization of the reaction norm for all genes in an organism's transcriptome across an ecologically-relevant thermal gradient. Rather than focusing on the identity of individual genes with differential expression, we compare overall patterns in the reactionomes of two closely-related temperate-zone species to gain insight into how spatially divergent selection has changed patterns of thermally-responsive gene expression. We apply this approach in a comparative study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996], which have more southern and northern distributions, respectively, and diverged approximately .... years ago (bernice??). Temperature is a potentially strong selective force on ground-nesting *Aphaenogaster* colonies, which must tolerate seasonally freezing winters and hot summers. On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@lubertazzi2012]. We collected RNAseq data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina (cite) paired-end short read data, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis (Supplemental Information).

As *A. carolinensis* lives in a climate that is hotter and has a longer growing season than that of *A. picea* (Supplemental Information), we expected patterns of gene expression in response to thermal extremes to differ between these species in a number of ways. First, the number of genes showing plasticity of expression to thermal extremes may differ between the species. Finding more genes are up-regulated to high temperatures in the colder-climate species *A. picea* would indicate that this species relies on activation of the cellular stress response [@kultz2005] and other protective genes to a greater extent than *A. carolinensis*, while the opposite would be true for cold temperatures. Conversely, finding that more genes are down-regulated in response to thermal extremes would indicate that a species relies to a greater extent upon proteome stability [@ghosh2010] and behavioral mechanisms [@sunday2014] to tolerate thermal stress.

@@

Importantly, none of the above hypotheses depend on the functional annotation of the responsive genes, which can be inaccurate in non-model systems [@primmer2013; @skunca2012]. However, insight on the molecular toolkit involved in adaptation to thermal extremes can be gained through circumspect examination of the functional annotation of responsive genes using gene set enrichment analysis. For example, we expect genes in the cellular stress response [@kultz2005] to be up-regulated in response to thermal extremes, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated in response to thermal extremes [@somero2011].

@@

## Species comparisons in the number of thermally-responsive genes

The combined *Aphaenogaster* transcriptome assembly contained 99,861 transcripts, hereafter referred to as genes, of which 47% had a best hit to an organism in the class Insecta.  Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` genes with expression patterns that were thermally-responsive at 5% False Discovery Rate. Thus, at a minimum `r round(9100/99861, 2)*100`% of genes are thermally-responsive. However, the closest related ant species with sequenced genomes have only about 17,00 genes [@gadau2012]. Assuming that the number of genes is similar for *Aphaenogaster*, this suggests that up to 50% of the genome (9,100 / 17,000) could be thermally-responsive.

```{r mosaic_plot, echo=FALSE, results='asis', fig.cap = "Mosaic plot of the number of thermally-responsive genes by expression type. The width of the columns is proportional to the number of genes for each expression type for *A. carolinensis* while the height of the rows is proportional to the number of genes for each expression type for *A. picea.* The rows are colored by expression type for *A. picea*. The total number of genes for two categories (*A. picea* Low - *A. carolinensis* Low; *A. picea* NotResp - *A. carolinensis* Bimodal) is labeled in two boxes for perspective."}
mosaic.colors <- brewer.pal(5, "Blues")
par(mar = c(5.1, 3.1, 2.1, 7.5))
mosaicplot(tt2, ylab = expression(italic("A. picea")), xlab = "", main = "", color = mosaic.colors, cex.axis = 0.01)
labels <- c("Low", "Intermediate", "High", "Bimodal", "NotResp")
text(c(.1,.5,.75,.85,.99), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE, font = 1)
text(.5, par("usr")[4]+.1, "A. carolinensis", xpd = TRUE, font = 3)
text(.15, .7, "2,594")
text(.85, .03, "223", cex = .75)
legend(1.05, .9, fill = mosaic.colors, c("Low", "Intermediate", "High", "Bimodal", "NotResp"), xpd=TRUE)
```

For each species, the thermally-responsive genes were grouped into five expression categories based on their reaction norms: genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (*NotResp*). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C. Of the thermally-responsive genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` had an expression pattern that differs between the two species. While some of these were simply responsive in only one species, `r colSums(tt2)[which(colnames(tt2) == "NotResp")]` genes that were responsive in *A. carolinensis* but not *A. picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` that were responsive in *A. picea* but not *A. carolinensis*, most of these were genes that changed from one expression category to another between the species (e.g. the number of genes in each expression category differed significantly between species; $\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig. 1). In both species, the majority of genes were in the **Low** category (Fig. 1). For *A. carolinensis*, this was followed by **Intermediate**, **High** and then **Bimodal**, whereas for *A. picea* the second largest category was **Bimodal** followed by **High** and finally **Intermediate** (Fig. 1).

## Sensitivity of thermally-responsive genes to changes in temperature

To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the temperature of gene activation for **High** and **Low** genes. The **High** genes in *A. carolinensis* were activated on average at `r T_on.high.ttest$estimate[1]`°C compared to `r T_on.high.ttest$estimate[2]`°C for *A. picea* (*t*~df=`r T_on.high.ttest$parameter`~ = `r T_on.high.ttest$statistic`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For genes with increased expression at **Low** temperatures, there was no difference in the mean temperature of gene activation (*t*~df=`r T_on.low.ttest$parameter`~ = `r T_on.low.ttest$statistic`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2).

For **Intermediate** and **Bimodal** genes, genes had two inflection points so it was not possible to calculate a single temperature of gene activation. Instead, we calculated the variance of 1,000 random temperatures, weighted by expression at that temperature. For **Intermediate** genes, a greater variance would be indicative of less sensitivity to changes in temperature. For **Bimodal** genes, a greater variance would be indicative of a more sensitivity to changes in temperature (convex reaction norm). We found that **Intermediate** genes from *A. carolinensis* had a greater variance of expression than those from *A. picea* (*t*~df=`r round(t.varint$parameter, 0)`~ = `r round(t.varint$statistic, 2)`, *P* < `r max(t.varint$p.value, 0.001)`, 95% CI difference `r paste(round(t.varint$conf.int[1], 3), "--", round(t.varint$conf.int[2], 3))`), indicating greater inertia of gene expression to changes in temperature. However, we found no difference between species in the variance of the reaction norms (*t*~df=`r round(t.varbim$parameter, 0)`~ = `r round(t.varbim$statistic, 2)`, *P* < `r round(t.varbim$p.value, 2)`) for the **Bimodal** genes.

@@

We looked for evidence of genetic assimilation (the process by which physiologically-relevant genes change from being environmentally-induced to basally-expressed )by comparing gene expression at intermediate temperatures for genes that were up-regulated at low temperatures in *A. carolinensis* or high temperatures in *A. picea*. Contrary to expectations, for **Low** genes in *A. carolinensis* gene expression at intermediate temperatures was lower in *A. picea* than *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`) . Moreover, for **High** genes in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`) and **Intermediate** genes in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`), gene expression at intermediate temperatures was always higher in *A. carolinensis*.

@@

As the molecular pathways involved in thermal response to cold and hot temperatures differ, we used gene set enrichment analysis on each expression category separately. First, we looked at the genes shared in each expression category (e.g. the intersection) for each species (Table S1). Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched (*P* < 0.01) terms for **High** genes. Clustering of terms by semantic similarity yielded 3 clusters, with 5 of the 7 terms being involved in "regulation of biosynthetic/metabolic process". For **Low** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` GO terms were enriched. These terms grouped into 15 clusters, including "protein modification", "protein metabolic process", "lipid metabolic process", "regulation of signaling", "protein transport/localization", "detection of stimulus", and "chondroitin sulfate biosynthesis". For **Bimodal** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched. These terms grouped into 5 clusters, including "biological regulation", "regulation of cellular biosynthesis", "gene expression", "signaling", and "protein modification process". For **Intermediate** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched. These terms grouped into 3 clusters, including "lipid biosynthesis/metabolism", "glycoprotein biosynthesis/metabolism", and "DNA organization". Overall, GSEA indicates that molecular responses to thermal extremes involve large-scale cellular and organizational changes, though the molecular responses to high temperatures are more limited than those to low temperatures.

We further used GSEA for the genes that shifted from being shut-down at extreme temperatures, **Intermediate**, in *A. carolinensis* to having increased expression in **Low** temperatures in *A. picea*. This group included `r nrow(Ar.int.A22.low.gsea)` GO terms that grouped into 12 clusters, including "protein metabolic process", "glycosylation" "transport", "lipid metabolic process", "GTP metabolism", "positive regulation of cellular process" and "organelle organization". For **Intermediate** genes in *A. carolinensis* that shifted **Bimodal** in *A. picea*, there were `r nrow(Ar.int.A22.bim.gsea)` enriched GO terms including "phospholipid catabolic process", "L-amino acid transport", "macromolecule metabolic process", "gluatmate receptor signaling pathway" and "histone modification". These results indicate that in response to thermal extremes, the warmer-climate species, *A. carolinensis*, shuts down metabolism to a greater extent than the colder-climate species, *A. picea*, which increases metabolism.

@@

Thermal tolerance is a key trait for determining the survival of individual organisms, the persistence of populations and the geographic limits of species. In this study, we applied a novel regression design to characterize the thermal reactionome of parapatric ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis*, found in colder and warmer environments, respectively, that are known to differ in their thermal tolerances [@warren2013]. To date, transcriptomic studies have used differential expression analysis between a few different conditions or environments (but see [@hodgins-davis2012]). By using a regression-approach, we are able to quantitatively describe changes in gene expression, rather than simply report the numbers of differentially-expressed genes. Specifically, we found that at least 10%, and potentially 50%, of the genes in these species were thermally-responsive.

The patterns of gene expression between species differed substantially, with more genes down-regulated in *A. carolinensis* and more up-regulated in *A. picea* in response to thermal extremes (Fig. 1, 3). The results of the gene set enrichment analysis give an indication of what might drive this pattern. The genes that are shut-down by *A. carolinensis* in response to cold temperatures, but up-regulated by *A. picea* include gene ontology clusters involved in metabolism, transport and regulation. Thus, *A. carolinensis*, the more southerly of our two species, shuts-down metabolism to a lesser extent than *A. picea* and so must rely on other mechanisms such as behavior (e.g. retreat into its nest) or proteome stability to tolerate thermal stress. In contrast, *A. picea* increases metabolism to a greater extent than *A. carolinensis* with more extreme temperatures, potentially to mount a greater molecular response to thermal stress. Consistent with this hypothesis, we find that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A. picea*, but only 4 do in *A. carolinensis*. This result supports our expectation that *A. picea* would rely to a greater extent upon activation of the cellular stress response and other protective genes to a greater extent than *A. carolinensis*. In general this contributes to evidence from other systems [@barshis2013]  that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance. ...Sara - something about enzyme efficiency tradeoffs???

@@

Given the differences in the number of thermally-responsive genes between the two species (Fig. 1), it is worth noting the similarity in expression patterns between species. There are 3-4\times more genes upregulated at cold than high temperatures in both species (Fig. 3). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures. At least among *Drosophila* taxa, these limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012]. In this light, for *A. picea* and *A. carolinensis*, our results suggest that adaptability to increased temperatures is limited, and these species risk replacement by more heat-tolerant competitors [@rodriguez-cabal2012].

In contrast to expectations that adaptation occurs via changes from plastically-induced to constitutive expression (e.g. genetic assimilation), we found lower basal gene expression in *A. picea* for genes that were up-regulated in response to cold temperatures (**Low** genes) in *A. carolinensis*. We also observed greater basal expression in *A. carolinensis* for **High** genes in *A. picea* as well as for genes that were **Intermediate** in both species. A potential explanation for the consistently higher basal gene expression in the warmer-environment species, *A. carolinensis*, is that this species has a higher metabolic rate and thus requires greater gene expression.
